Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data

### 2021-01-25 manuscript/manuscript.tex
Revisions following round 2 of review

Vertical drift can occur quite unpredictably, even following good quality calibration, and it is likely caused by spatial phenomena such as degraded eye tracker calibration at the extremes of the screen, or temporal phenomena such as subtle movements in head position or pupil dilation, which can be difficult to control for, even in a laboratory setting. Such sources of measurement error are often exacerbated in the context of multiline reading because, in comparison to single words or sentences, passages of text are distributed over a larger portion of the screen, including areas where general calibration may be worse, and they take longer to read, during which time calibration may begin to degrade. There are also less frequent opportunities to recalibrate the device during passage reading, since this can only be performed between trials or pages and not during the reading of a passage.

@@

To be clear, the algorithms we consider in this paper are restricted to one specific problem. Firstly, we only consider algorithms designed for the ordinary reading of passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are quite different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014}. Similarly, the reading of source code has received some attention, but the affordances and constraints in this domain are quite different from ordinary linguistic reading \parencite[see e.g.,][]{Palmer:2016,Nussli:2011}. Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Sibert:2000}. Thirdly, we only consider fully automated algorithms that do not require human supervision.

@@

The algorithm then enters the next ``phase'' of the process, in which the criteria are slightly relaxed, allowing more mergers to occur. These phases could in principle be defined by the user, but we follow the four-phase model reported by \textcite{Spakov:2019}, which effectively builds a set of heuristics into the algorithm. In Phase~1, the first and second sequences must each contain a minimum of 3~fixations to be considered for merging; in Phase~2, only the second sequence must contain a minimum of 3~fixations; in Phase~3, there is no minimum number of fixations; and in Phase~4, the gradient and regression error criteria are also entirely removed. Of course, as soon as the number of sequences is reduced to $m$ the algorithm exits the merge process, so not all four phases will necessarily be required. Finally, the set of $m$ sequences is matched to the set of text lines in positional order: The sequence with the smallest mean \textit{y}~value is mapped to line one and so forth.

@@

We analyze the performance characteristics of the algorithms in three ways. Firstly, we look at how the algorithms fare against the gold standard manual correction; secondly, we look at how the algorithms perform in comparison to no drift correction at all; and thirdly, we look at how the algorithms relate to each other, regardless of their accuracy.

@@

It is also worth considering what proportion of trials are likely to be usable following the application of a particular method. Fig.~\ref{fig10} reports the proportion of trials that obtained accuracy of 90\%, 95\%, and 99\% by algorithm. If you are willing to accept relatively low accuracy at the trial level (e.g., 90\%, Fig~\ref{fig10}A), then \texttt{cluster}, \texttt{merge}, and \texttt{stretch} will provide the best performance---a large proportion of trials will meet this criterion. In comparison, if you have more stringent accuracy requirements at the trial level (e.g., 99\%, Fig~\ref{fig10}C), then \texttt{segment}, \texttt{split}, and \texttt{warp} are likely to provide better performance. Of course, the cost of a more stringent accuracy criterion is that fewer trials will be usable overall, and it is not possible to know which trials have low accuracy in the absence of manual correction data. This highlights the fact that it is currently not possible to confidently achieve a high level of accuracy in a high proportion of trials, so researchers may still need to invest a significant amount of time if a high level of accuracy is demanded.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig10_double_column.eps}}
	\vspace*{2pt}
	\caption{Proportion of trials that reached at least 90\% (A), 95\% (B), and 99\% (C) accuracy by algorithm.}
	\label{fig10}
	\end{figure}

@@

A second, perhaps more useful, way to measure the performance of the algorithms is in terms of how much of an improvement an algorithm provides in comparison to applying no drift collection at all. To estimate this, we first need to define a baseline level of accuracy. As mentioned previously, the \texttt{attach} algorithm essentially corresponds to a standard eye tracking analysis; it is equivalent to drawing maximal, nonoverlapping bounding boxes around the words in a passage and then mapping fixations to whichever bounding box they fall into (as would be the case in a standard analysis of eye tracking data using the widely adopted Area-of-Interest paradigm). Therefore, we can estimate the potential improvement that a given algorithm offers by comparing its accuracy to the accuracy of the \texttt{attach} algorithm.

>>>>

Vertical drift can occur quite unpredictably, even following good quality calibration, and it is likely caused by spatial phenomena such as degraded eye tracker calibration at the corners of the screen, or temporal phenomena such as subtle movements in head position or pupil dilation, which can be difficult to control for, even in a laboratory setting. Such sources of measurement error are often exacerbated in the context of multiline reading because, in comparison to single words or sentences, passages of text are distributed over a larger portion of the screen, including areas where general calibration may be worse, and they take longer to read, during which time calibration may begin to degrade. There are also less frequent opportunities to recalibrate the device during passage reading, since this can only be performed between trials or pages and not during the reading of a passage.

@@

To be clear, the algorithms we consider in this paper are restricted to one specific problem. Firstly, we only consider algorithms designed for the ordinary reading of passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are quite different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014,Vadillo:2015}. Similarly, the reading of source code has received some attention, but the affordances and constraints in this domain are quite different from ordinary linguistic reading \parencite[see e.g.,][]{Palmer:2016,Nussli:2011}. Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Sibert:2000}. Thirdly, we only consider fully automated algorithms that do not require human supervision.

@@

The algorithm then enters the next ``phase'' of the process, in which the criteria are slightly relaxed, allowing more mergers to occur. These phases could in principle be defined by the user, but we follow the four-phase model reported by \textcite{Spakov:2019}, which effectively builds a set of heuristics into the algorithm. In Phase~1, the first and second sequences must each contain a minimum of three fixations to be considered for merging; in Phase~2, only the second sequence must contain a minimum of three fixations; in Phase~3, there is no minimum number of fixations; and in Phase~4, the gradient and regression error criteria are also entirely removed. Of course, as soon as the number of sequences is reduced to $m$ the algorithm exits the merge process, so not all four phases will necessarily be required. Finally, the set of $m$ sequences is matched to the set of text lines in positional order: The sequence with the smallest mean \textit{y}~value is mapped to line one and so forth.

@@

We analyze the performance characteristics of the algorithms in four ways. Firstly, we look at how the algorithms fare against the gold standard manual correction; secondly, we look at what proportion of trials are likely to be usable following drift correction; thirdly, we look at how the algorithms perform in comparison to using no drift correction at all; and finally, we look at how the algorithms relate to each other, regardless of their accuracy.

@@

\subsubsection{Proportion of Corrections Likely to be Usable}

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig10_double_column.eps}}
	\vspace*{2pt}
	\caption{Proportion of trials that surpassed (A) 90\%, (B) 95\%, and (C) 99\% accuracy, and (D) the proportion of corrections deemed acceptable by two human raters. The dark and light bars represent the adult and child datasets respectively.}
	\label{fig10}
	\end{figure}

Fig.~\ref{fig10} reports the proportion of corrections that surpassed an accuracy level of 90\%, 95\%, and 99\% by algorithm. If you are willing to accept relatively low accuracy at the trial level (e.g., 90\%, Fig.~\ref{fig10}A), then \texttt{cluster}, \texttt{merge}, and \texttt{stretch} will provide the best performance---a large proportion of corrections will meet this criterion. In comparison, if you have more stringent accuracy requirements at the trial level (e.g., 99\%, Fig.~\ref{fig10}C), then \texttt{segment}, \texttt{split}, and \texttt{warp} are likely to provide better performance. Of course, the cost of a more stringent accuracy criterion is that fewer corrections will be usable overall, and it is not possible to know which trials have low accuracy in the absence of manual correction data. This highlights the fact that it is currently not possible to confidently achieve a high level of accuracy in a high proportion of trials, so researchers may still need to invest a significant amount of time if a high level of accuracy is demanded.

To estimate how usable the corrections are likely to be to a typical researcher, we performed a more subjective analysis of their quality. All 480 algorithmic corrections were presented blind and in random order to two raters (JWC and VNP), who independently classified every correction as either ``acceptable'' or ``needs more work.'' The raters did not discuss in advance what criteria they would use to make these judgments, but agreement was nevertheless very high at 94\%. In addition to the overall number of errors, the raters weighed up other factors, such as how the errors were distributed over the passage, how challenging the input data seemed to be, and what effect the errors might have for downstream analyses. The results, which are shown in Fig.~\ref{fig10}D, suggest that \texttt{cluster}, \texttt{merge}, and \texttt{stretch} are likely to produce very satisfactory results on adult data.

@@

Another way to measure performance is in terms of how much of an improvement an algorithm provides in comparison to applying no drift collection at all. To estimate this, we first need to define a baseline level of accuracy. As mentioned previously, the \texttt{attach} algorithm essentially corresponds to a standard eye tracking analysis; it is equivalent to drawing maximal, nonoverlapping bounding boxes around the words in a passage and then mapping fixations to whichever bounding box they fall into (as would be the case in a standard analysis of eye tracking data using the widely adopted Area-of-Interest paradigm). Therefore, we can estimate the potential improvement that a given algorithm offers by comparing its accuracy to the accuracy of the \texttt{attach} algorithm.
